Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-10-11 manu-draft.qmd
added mcb comments

The health of estuarine systems and coastal habitats is tightly linked to land use and management of the watershed [@Yoskowitz15]. Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18]. Pervasive stressors can negatively impact the services provided by these habitats. Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14]. Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19]. Management of coastal habitats requires an understanding of how stressors impact these resources, while balancing competing societal uses.

@@

Tampa Bay is a large open water estuary (surface area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap). The watershed covers approximately 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. The climate is subtropical and within the 2020 ecotone for mangrove and salt marsh habitats. Naive habitats in the watershed include pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.

@@

Subtidal habitats included hard bottom [@Jaap90; @Ash05; @Kaufman17; @CSA19], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Intertidal habitats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15]. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below). This 5-foot contour is an area of intense urban development and is expected to be impacted by sea-level rise based on current estimates. Since 1946, the St. Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day. Projections to the year 2100 suggest sea levels can increase between 58 and 259 cm in the region [@Burke19].

@@

Coverage targets for habitat types and opportunity areas for restoration were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the watershed were synthesized to inform on a practical and feasible level of effort that could be conducted by restoration practitioners in the future. The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for native habitat coverage (hectares). Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise. As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.

@@

Emergent tidal and freshwater wetlands in Tampa Bay have also experienced dramatic changes over the last three decades. Dual pressures from sea-level rise and climate change have impacted tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space. Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12; @Osland22]. As such, the identified targets and goals for mangroves indicate protection of these habitats, without the need for additional restoration. However, mangroves are expected to continue colonization of the intertidal zone, contributing to additional losses of salt marshes and salt barrens. The reservation areas identified in @fig-oppmap represent critical remaining areas in the intertidal zone that could be protected to prevent additional losses of tidal wetlands. Likewise, gains in non-forested freshwater wetlands are a reflection of 1) the effectiveness of state and federal wetland regulatory programs, and 2) the cumulative gains from publicly-funded habitat restoration projects. A similar approach could be used for salt marshes or salt barrens.

The decrease in native uplands is the result of continued development in the Tampa Bay watershed (@fig-chgdatalluv) and lack of regulatory protection of these habitats. Attaining the target and goal will require restoration of upland habitats on existing conservation lands (i.e., restoration potential in @fig-restmap) and new conservation lands to offset the continued loss of these habitats to development. The long-term conversion of restorable areas to developed lands (@fig-chgdatalluv) presents additional challenges for restoration of native uplands. Amendments to existing planning, zoning, and land development policies will be needed to address these issues. Although federal and state regulations for endangered species provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods). Reductions in native uplands will likely continue unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development ordinances to maintain a defined extent of these habitats within a rapidly urbanizing coastal watershed.

@@

Creative approaches may be required in areas impacted by sea-level rise if land acquisition is not possible. These approaches are necessary to accommodate future landward migration of tidal wetlands or the protection of coastal uplands, while also preventing or minimizing development activities that can inhibit landward habitat shifts. Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline. However, rolling easements may be an alternative approach whereby protected areas are allowed to "roll" landward with expected changes in sea-level rise. Rolling easements could disincentivize more intense urban development of low-lying coastal uplands in less developed agricultural or recreational land uses. Landowners could maintain current economic uses with a rolling easement, while reserving such lands to accommodate landward habitat migration.

Finally, wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs could serve as more focused and directed restoration mechanisms to help achieve watershed-wide goals. Mitigation activities have historically been conducted independent of watershed-level planning and monitoring processes. This disconnect has contributed to fragmented implementation and habitat function and inconsistent compliance monitoring of mitigation projects, including historically poor documentation of wetland losses and gains in the Tampa Bay watershed. However, if properly focused and coordinated, compensatory mitigation activities could significantly contribute to the attainment of restoration targets and goals for the region.

@@

Identifying restoration priorities was data intensive and would not have been possible without the resources available for the region. The workflow for identifying priorities required detailed and spatially-explicit datasets specific to the Tampa Bay watershed. Long-term datasets describing land use and cover and the extent of subtidal habitats were necessary to categorize current extent and past trends. Similarly, supporting datasets included those that described existing and proposed conservation areas, soils, past restoration activities, and relevant spatial boundaries (i.e., watershed and coastline). Many of these datasets are available outside of Tampa Bay, although temporal and spatial resolutions may limit application to other areas. Despite the region’s data richness, limitations still exist in classifying restoration opportunities based on the spatial-scale of the LULC datasets. Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets. Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types. For example, inventories of hard bottom habitats, living shorelines, tidal tributaries, and artificial reefs were created through special studies or were available only as current estimates from regional entities. Tracking progress towards these habitat targets and goals is heavily reliant on regular updates to these datasets, as well as routine land use and cover map updates.

@@

![Spatial analysis workflow used to identify opportunity areas and restoration potential in the Tampa Bay watershed. Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas and restoration potential.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}

>>>>

The health of estuarine systems and coastal habitats is tightly linked to land use and management of the watershed [@Yoskowitz15]. Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18]. Pervasive stressors can negatively impact the services provided by these habitats. Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14]. Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19]. Management of coastal habitats requires an understanding of how stressors influence these resources, while balancing competing societal uses.

@@

Tampa Bay is a large open water estuary (surface area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap). The watershed covers approximately 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. The climate is subtropical and within the 2020 ecotone for mangrove and salt marsh habitats. Native habitats in the watershed include pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.

@@

Subtidal habitats included hard bottom [@Jaap90; @Ash05; @Kaufman17; @CSA19], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Intertidal habitats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15]. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below). This 5-foot contour is an area of intense urban development and is expected to be affected by sea-level rise based on current estimates. Since 1946, the St. Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day. Projections to the year 2100 suggest sea levels can increase between 58 and 259 cm in the region [@Burke19].

@@

Coverage targets for habitat types and opportunity areas for restoration were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the watershed were synthesized to inform on a practical and feasible level of effort that could be conducted by restoration practitioners in the future. The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for native habitat coverage (hectares). Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be affected by sea-level rise. As such, the approach identifies reasonable goals and targets based on past trends and provides spatially explicit information that identifies where restoration practitioners could prioritize projects based on opportunities within their respective jurisdictions.

@@

Emergent tidal and freshwater wetlands in Tampa Bay have also experienced dramatic changes over the last three decades. Dual pressures from sea-level rise and changes in the length of freeze-free season have affected tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space. Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12; @Osland22]. Anthropogenic water withdrawals have also reduced freshwater flows reaching tidal marshes, contributing to reductions in coverage of key species (e.g., *Juncus roemerianus*) in place of mangroves [@Raabe12].  As such, the identified targets and goals for mangroves indicate protection of these habitats, without the need for additional restoration. However, mangroves are expected to continue colonization of the intertidal zone, contributing to additional losses of salt marshes and salt barrens. The reservation areas identified in @fig-oppmap represent critical remaining areas in the intertidal zone that could be protected to prevent additional losses of tidal wetlands. Likewise, gains in non-forested freshwater wetlands are a reflection of 1) constructed stormwater ponds required by state and federal regulatory programs, and 2) the cumulative gains from publicly-funded habitat restoration projects. Creative restoration approaches (e.g., habitat acquisition and optimal management of freshwater flows) that address the likely expansion of mangroves at the expense of salt marshes and salt barrens will be required to meet the targets and goals for these habitats.

The decrease in native uplands is the result of continued development in the Tampa Bay watershed (@fig-chgdatalluv) and lack of regulatory protection of these habitats. Attaining the target and goal will require restoration of upland habitats on existing conservation lands (i.e., restoration potential in @fig-restmap) and new conservation lands to offset the continued loss of these habitats to development. The long-term conversion of restorable areas to developed lands (@fig-chgdatalluv) presents additional challenges for restoration of native uplands. Additional education about best practices in land development, market-based incentives, and amendments to existing planning, zoning, and land development regulations will be needed to address these issues. Although federal and state regulations for endangered species provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods). Voluntary approaches to low impact urban or suburban development may also gain traction among developers as more viable methods for land conversion that minimize impacts to natural resources while increasing quality of life [@Jones09].  Education and outreach activities that target land developers to raise awareness of the benefits of alternative practices are critical in this effort.  Market-based approaches to mitigating urban sprawl may also be practical [e.g., conservation subdvisions, @Mohamed06] given the estimated economic gains relative to conventional approaches. Regardless, reductions in native uplands will likely continue in the short-term unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development regulations to maintain a defined extent of these habitats within a rapidly urbanizing coastal watershed.

@@

Creative approaches may be required in areas affected by sea-level rise if land acquisition is not possible. These approaches are necessary to accommodate future landward migration of tidal wetlands or the protection of coastal uplands, while also reducing risks to built infrastructure that, when inappropriately sited, can inhibit landward habitat shifts. Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline. However, rolling easements may be an alternative approach whereby protected areas are allowed to "roll" landward with expected changes in sea-level rise. Rolling easements could disincentivize more intense urban development of low-lying coastal uplands in less developed agricultural or recreational land uses. Landowners could maintain current economic uses with a rolling easement, while reserving such lands to accommodate future landward habitat migration. These approaches also offer risk-reduction to built infrastructure by offering increased protection from potential affects of sea-level rise and other coastal stressors (e.g., storm surge).

Finally, wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs could serve as more directed restoration mechanisms to help achieve watershed-wide goals. Mitigation activities have historically been conducted independent of watershed-level planning and monitoring processes. This disconnect has contributed to fragmented implementation, marginal habitat function, and inconsistent compliance monitoring of mitigation projects, including historically poor documentation of wetland losses and gains in the Tampa Bay watershed. However, if properly focused and coordinated, compensatory mitigation activities could significantly contribute to the attainment of restoration targets and goals for the region.

@@

Identifying restoration priorities was data intensive and would not have been possible without the resources available for the region. The workflow for identifying priorities required detailed and spatially-explicit datasets specific to the Tampa Bay watershed. Long-term datasets describing land use and cover and the extent of subtidal habitats were necessary to categorize current extent and past trends. Similarly, supporting datasets included those that described existing and proposed conservation areas, soils, past restoration activities, and relevant spatial boundaries (i.e., watershed and coastline). Many of these datasets are available outside of Tampa Bay, although temporal and spatial resolutions may limit application to other areas. Despite the region’s data richness, limitations still exist in classifying restoration opportunities based on the spatial-scale of the LULC datasets. Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets, which could expose additional opportunities in urban or suburban areas (e.g., improving stormwater infrastructure). Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types. For example, inventories of hard bottom habitats, living shorelines, tidal tributaries, and artificial reefs were created through special studies or were available only as current estimates from regional entities. Tracking progress towards these habitat targets and goals is heavily reliant on regular updates to these datasets, as well as routine land use and cover map updates.

@@

![Spatial analysis workflow used to identify opportunity areas (existing conservation native, proposed conservation native, existing conservation restorable, proposed conservation restorable, reservation native, reservation restorable) and restoration potential (coastal uplands, uplands, freshwater wetlands, tidal wetlands) in the Tampa Bay watershed. Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas and restoration potential.  The approach was applied to both the intertidal and supratidal strata of the watershed.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}
